Experiences with Planning for Natural Language Generation

### 2009-06-02 trunk/Papers/ci-spark/abstract.tex
changed to official CI style
changes to abstract, section1, and section 2

%%%
%%% Abstract
%%%
\begin{abstract}
  Natural language generation (NLG) is a field of computational
  linguistics that has a long tradition as an application area of
  planning systems. While this connection had not received much
  attention for a while, interest in it has been renewed recently in
  several publications. In this paper, we investigate the extent to
  which these new NLG systems profit from the recent advances in
  planner efficiency. Our findings are mixed: While modern planners do
  very well on solving the search problems that come up in our NLG
  experiments, their overall runtime is dominated by the grounding
  step they perform as preprocessing. Because this makes current
  off-the-shelf planners unusably slow for nontrivial NLG problem
  instances, we offer our domains as challenges for the planning
  community.
\end{abstract}

\bigskip\noindent
\textbf{Keywords:} natural language generation, planning

>>>>

\begin{abstract}
  Natural language generation (NLG) is a major subfield of
  computational linguistics with a long tradition as an application
  area of automated planning systems. While the connection between
  NLG and planning has not received much attention in current
  mainstream NLG research, several recent publications have sparked
  a renewed interest in this area. In this paper, we investigate the
  extent to which these new NLG approaches profit from the advances
  in planner expressiveness and efficiency. Our findings are mixed.
  While modern planners can readily handle the search problems that
  arise in our NLG experiments, their overall runtime is often
  dominated by the grounding step they perform as preprocessing.
  Furthermore, small changes in the structure of a domain can
  significantly shift the balance between search and preprocessing.
  Overall, our experiments show that the off-the-shelf planners we
  tested are unusably slow for nontrivial NLG problem instances. As
  a result, we offer our domains and experiences as challenges for
  the planning community.
\end{abstract}

\begin{keywords}
  natural language generation, planning
\end{keywords}

### 2009-06-02 trunk/Papers/ci-spark/introduction.tex
changes to abstract, section1, and section 2

Treating generation as planning has a long history in NLG, ranging
from the initial attempts of the field to utilise early planning
approaches \citep{perrault80,appelt:planning,hovy88,young94dpocl}, to
a recent surge of research
\citep{Steedman-Petrick:07,KolSto07,brenner08:_contin_multiag_plann_approac_to_situat_dialog,benotti08b}
seeking to capitalise on the improvements modern planners offer in
terms of efficiency and expressiveness. This paper attempts to assess
the usefulness of current planning techniques to NLG by investigating
some representative generation problems, and evaluating whether
automated planning has advanced to the point that it can provide
solutions to such NLG applications---applications that are not
currently being investigated by mainstream planning research.

@@

Second, we evaluate the performance of current off-the-shelf planners
in the planning domains into which these two generation problems
translate. Among the planners we test, we explore the efficiency of FF
\citep{HoffmannNebel01}---a planner that has arguably had the greatest
impact on recent approaches to deterministic planning---and some of its
descendants, such as SGPLAN
\citep{hsu06:_new_featur_in_sgplan_for}. All of the planners we test
are freely available, support an expressive subset of the Planning
Domain Definition Language (PDDL; McDermott et al.~\citeyear{PDDL}),
and have been successful on both standard planning benchmarks and the
problems of the International Planning Competition (IPC).\footnote{See
  \url{http://ipc.icaps-conference.org/} for information about recent
  and past editions of the IPC. A good overview of the deterministic
  track of the 2004 instance of the IPC can be found in
  \citep{Hoffmann-Edelkamp:05}.}  Using these planners---and an ad-hoc
Java implementation of GraphPlan \citep{Blum1997} which serves as a
baseline for certain experiments---we perform a series of tests on a
range of problem instances in our NLG domains.

Overall, our findings are mixed. On the one hand, we demonstrate that
some planners can readily handle the \emph{search} problems that arise
in our testing domains on realistic inputs, which is promising given
the challenging nature of these tasks (e.g., the sentence generation
task is NP-complete; see \citealt{KolStr02}). On the other hand, these
same planners spend tremendous amounts of time on \emph{preprocessing}
to analyse the problem domain in support of the search. On many of our
problem instances, the preprocessing time overshadows the search time.
For instance, FF spends 90\% of its runtime in the sentence generation
domain on grounding out actions and literals. As a consequence, we
conclude that the off-the-shelf planners we investigated are too slow
to be useful in real NLG applications.  We hope that our results will
spark some interest for more efficient implementations of planning
preprocessors, and offer our NLG domains and experiences as challenges
for the planning community.

The remainder of this paper is structured as follows. In
Section~\ref{sec:nlg-as-planning}, we introduce the idea of NLG as
planning and briefly review the relevant literature. In
Section~\ref{sec:domains}, we describe a set of planning problems
associated with two NLG tasks, namely sentence planning and situated
instruction generation. In Section~\ref{sec:experiments}, we report on
the results of our experiments with these planning problems.  In
Section~\ref{sec:discussion} we discuss our results and overall
experiences, and conclude in Section~\ref{sec:conclusion}.

>>>>

Treating generation as planning has a long history in NLG, ranging
from the initial attempts of the field to utilise early planning
approaches \citep{perrault80,appelt:planning,hovy88,young94dpocl}, to
a recent surge of research
\citep{Steedman-Petrick:07,KolSto07,brenner08:_contin_multiag_plann_approac_to_situat_dialog,benotti08b}
seeking to capitalise on the improvements modern planners offer in
terms of efficiency and expressiveness. This paper attempts to assess
the usefulness of current planning techniques to NLG by investigating
some representative generation problems, and by evaluating whether
automated planning has advanced to the point that it can provide
solutions to such NLG applications---applications that are not
currently being investigated by mainstream planning research.

@@

Second, we evaluate the performance of several off-the-shelf planners
on the planning domains into which these two generation problems
translate. Among the planners we test, we explore the efficiency of FF
\citep{HoffmannNebel01}---a planner that has arguably had the greatest
impact on recent approaches to deterministic planning---and some of its
descendants, such as SGPLAN
\citep{hsu06:_new_featur_in_sgplan_for}. All of the planners we test
are freely available, support an expressive subset of the Planning
Domain Definition Language (PDDL; McDermott et al.~\citeyear{PDDL}),
and have been successful on both standard planning benchmarks and the
problems of the International Planning Competition (IPC).\footnote{See
  \url{http://ipc.icaps-conference.org/} for information about past
  editions of the IPC. Also see \citep{Hoffmann-Edelkamp:05} for a good
  overview of the deterministic track of the 2004 competition.}
Using these planners---together with an ad-hoc Java implementation of
GraphPlan \citep{Blum1997} serving as a baseline for certain
experiments---we perform a series of tests on a range of problem
instances in our NLG domains.

Overall, our findings are mixed. On the one hand, we demonstrate that
some planners can readily handle the \emph{search} problems that arise
in our testing domains on realistic inputs, which is promising given
the challenging nature of these tasks (e.g., the sentence generation
task is NP-complete; see \citealt{KolStr02}). On the other hand, these
same planners often spend tremendous amounts of time on
\emph{preprocessing} to analyse the problem domain in support of the
search. On many of our problem instances, the preprocessing time
overshadows the search time. (For instance, FF spends 90\% of its
runtime in the sentence generation domain on preprocessing.)
Furthermore, small changes in the structure of a planning domain can
dramatically shift the balance between preprocessing and search. As a
consequence, we are forced to conclude that the off-the-shelf planners
we investigated are generally too slow to be useful in real NLG
applications. It is also our hope, however, that these results will
spark an interest to improve the quality of planner
implementations---especially in the area of preprocessing
techniques---and to this end we offer our domains and experiences as
challenges for the planning community.

The remainder of this paper is structured as follows. In
Section~\ref{sec:nlg-as-planning}, we introduce the idea of NLG as
planning and briefly review the relevant literature. In
Section~\ref{sec:domains}, we describe a set of planning problems
associated with two NLG tasks: sentence planning and situated
instruction generation. In Section~\ref{sec:experiments}, we report on
our experiments with these planning problems.  In
Section~\ref{sec:discussion} we discuss our results and overall
experiences, and conclude in Section~\ref{sec:conclusion}.
